Applying the optimized ensemble histogram method to the square-well liquid

### 2015-01-07 papers/square-well-liquid/histogram-paper.tex
implemented vanilla wang landau, i.e. wang landau with the parameter values   given in the original WL paper, with one exception: I made the end condition less   stringent, because otherwise initialization took at least an hour on my computer. Tested on quipu. Took 8:07.89.
Tested on quipu. Took 0:30.86.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/periodic-ww130-ff30-N20-sample-rate}
  \caption{Round trip rates from different methods.}
\end{figure}

@@

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/periodic-ww130-ff30-N20-sample-rate}
  \caption{Inverse sampling rates from different methods.}
\end{figure}

@@

\subsection{Tracking transitions}

We can track the number of attempted transitions from each energy
state to each other energy state.  This statistic has the advantage of
being approximately independent of the weighting function used
(although it is highly dependent on the step size).  Thus we could
continue to refine this statistic, even as we are updating the weight
array.

The ratio of density of states at two energies should be proportional
to the ratio of upgoing and downgoing transition rates between those
two energies.  Thus these transitions give us information about
$\Delta D$ (where $D$ is the density of states).

Figure~\ref{fig:transitions} below shows the transition matrix for a
20-sphere system, computed during initialization using the Wang-Landau
algorithm.

How can we use this?

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/periodic-ww130-ff30-N20-transitions}
  \caption{A plot showing the probability of energy transitions (with
    no weighting) from given energy states.\label{fig:transitions}}
\end{figure}

\subsubsection{Generating $w$ from the transition matrix}

We should be able to generate a set of weighting functions directly
from the transition matrix.  We could do this either by optimizing for
a flat histogram, or by optimizing the round-trip rate.  Both would be
interesting to implement.

One natural approach would use linear algebra.  If the normalized
transition matrix is $T$, then the density of states is given by
\begin{equation}
  D = TD
\end{equation}
By solving this equation for $D$, we could find a good approximation
for $w$ to obtain a flat histogram.

I expect that we could use similar math to the optimized ensemble
folks to optimize the round-trip rate.

\subsubsection{Sampling the transition matrix (more) directly}

If our goal (in initialization) is to sample the transition matrix
(from which we compute the optimal weights), we could use a much more
direct approach.  When computing a histogram, we are forced to perform
a complete random walk.  However, the transition matrix can be
computed at a given energy \emph{without sampling any other energies}!

We can start at a given energy, sample the transition matrix at that
energy to our satisfaction, i.e. until we have observed transitions in
both directions a statistically significant number of times.  Then we
would proceed to the next energy and sample it, and so forth.  There
is a weakness in that we may not be able to sample all states at a
given energy without going to other energies.  But we could repeat
this procedure many times to accomplish this.  In a way, this
algorithm is reminiscent of Wang-Landau, which ``forces'' the
simulation to sample a whole slew of different energies, but in our
case we would be able to put strict bounds on how long it spends
sampling each energy, and could reasonably keep the code from wasting
huge amounts of time sampling the high-entropy states.
